Data Predictive Control with Random Forests:\\Theory and Applications\tnoteref{mytitlenote}

### 2017-09-26 paper/casestudyReal.tex
Modified introduction. First complete draft.

\section{REAL CASE STUDY: OPTIMAL HEATING SYSTEM SCHEDULING} In this section we show an application of DPC to a real case study. In particular, in the following we first describe the plant structure, then we provide both data-based and EnergyPlus models of such a plant to be used for control and performance comparison, and finally we show results of the closed-loop simulations.

@@

\begin{figure}[h!]
	\begin{center}
		\subfigure[Walls.]{
			\label{F:houseSection1}
			\centering
			\includegraphics[width=24pc]{figures/blocco_parete_rev01.eps}
		}
		\subfigure[Floor and roof.]{
			\label{F:houseSection2}
			\centering
			\includegraphics[width=24pc]{figures/blocco_solaio_rev01.eps}
		}
	\end{center}
	\caption{Cross-sections of the wood-cement blocks.}
	\captionsetup{justification=centering}
	\label{F:houseSection}
\end{figure}
The heating system of the building, that supplies the thermal energy required in winter season, is a hydronic system consisting of a vegetable biomass boiler, with a manual ON/OFF, a constant-flow pump station, and tubular steel radiators as shown in Figure \ref{F:housePlantScheme}. The standard biomass boiler has an efficiency of $83.5\%$ with $16.5\ kW$ of thermal power transferred to the water, without gas-flame modulation. The heat transfer fluid distribution is realized through a manifold circuit, that supplies the radiators placed in the various rooms of the house. The energy needs for the domestic hot water (DHW) are covered by the same boiler, coupled with a solar thermal plant. It is worth noting that, in this work, the thermal energy needed for the domestic hot water production is neglected.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=26pc]{figures/Schema_di_impianto.eps}
		\caption{Technological plant scheme of the use case.}
		\captionsetup{justification=centering}
		\label{F:housePlantScheme}
	\end{center}
\end{figure}
For a thorough knowledge of the building thermal behavior, an in-situ analysis was carried out. Temperature measuring devices, completely self-produced at the “G. Parolini Lab” of the University of L'Aquila \cite{Pantoli2017}, mainly based on an $ATmega 2560$ microcontroller and $DS18B20$ temperature probes (temperature range from $-55.0\degree C$ to $125.0\degree C$ with an accuracy of $\pm 0.5\degree C$), were employed to acquire the ambient temperatures of the four differently oriented rooms of the house as shown in Figure \ref{F:houseFloors}. The probes positions were chosen to optimize the data acquisition and to minimize the discomfort of the occupants in the house. Furthermore, as can be seen in Figures \ref{F:housePlantScheme} and \ref{F:houseFloors}, a commercial heat meter (temperature range from $10.0\degree C$ to $90.0\degree C$ with an accuracy of $\pm 0.05\degree C$) was installed downstream of the biomass boiler to measure the produced thermal energy. The turbine flowmeter was installed on the return pipe, while the two thermocouples were placed inside the delivery and return pipes, respectively. All the measuring devices have been set with a data acquisition rate equal to $10$ minutes, from March $11^{th}$ $2016$ to May $15^{th}$ $2016$.
\begin{figure}[h!]
	\begin{center}
		\subfigure[Ground floor.]{
			\label{F:houseGroundFloor}
			\centering
			\includegraphics[width=26pc]{figures/Arch_PT_e_sonde.eps}
		}
		\subfigure[Attic.]{
			\label{F:houseAttic}
			\centering
			\hspace{-1.6cm}
			\includegraphics[width=22.8pc]{figures/Arch_P1.eps}
		}
	\end{center}
	\caption{Layout of the house and probes placement. Legend: red circle for ambient temperature; orange circle for heat meter.}
	\captionsetup{justification=centering}
	\label{F:houseFloors}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=24pc]{figures/Render_rev01.eps}
		\caption{Virtual model of the use case.}
		\captionsetup{justification=centering}
		\label{F:houseVirtualModel}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\subfigure[Thermal energy.]{
			\label{F:houseComparisonExperimentalEnergy}
			\centering
			\includegraphics[width=24pc]{figures/Thermal_energy_rev01.eps}
		}
		\subfigure[Thermal power.]{
			\label{F:houseComparisonExperimentalPower}
			\centering
			\includegraphics[width=24pc]{figures/Thermal_power.eps}
		}
	\end{center}
	\caption{Comparison between numerical and experimental data.}
	\captionsetup{justification=centering}
	\label{F:houseComparisonExperimental}
\end{figure}
\subsection{Random forest models} For the closed-loop simulations with DPC we create $2$ different sets of models, $S_1 = \{\tP^{r}_{\mathrm{k+1|k}},\tT^{r_1}_{\mathrm{k+1|k}},\tT^{r_2}_{\mathrm{k+1|k}},\tT^{r_3}_{\mathrm{k+1|k}},\tT^{r_4}_{\mathrm{k+j|k}}\}$ and $S_2 = \{\tP_{\mathrm{k+j|k}},\tT^{1}_{\mathrm{k+j|k}},\tT^{2}_{\mathrm{k+j|k}},\tT^{3}_{\mathrm{k+j|k}},\\\tT^{4}_{\mathrm{k+j|k}},\ j = 1,\ldots,N\}$, using random forests. In each set we have a model for the power consumption of the house ($\tP^{r}$ and $\tP$) and $4$ models that describe the evolution of the room temperature in each of the $4$ rooms ($\tT^{r_i}$ and $\tT^i$). Models in $S_1$ are created using the classical random forests algorithm with all the features and are used as plant simulator of the house. For this reason, they are computed to give prediction only for time step $k+1$ and not over the whole horizon $N$. Models in $S_2$ are used as predictors over a horizon $N$ in the DPC algorithm and are then created using the methodology provided in Section \ref{SS:dpcrf}. The non-manipulated features in $\tX^d$ are the disturbance data (relative humidity, atmospheric pressure, outside air temperature, solar radiation, wind, time of the day and day of the week) and the states (temperature of the $4$ rooms). The manipulated feature in $\tX^c$ is the flow rate ($[m^3/h]$). All this features are used to create models in $S_1$ and the power model in $S_2$, while disturbance data, state temperature of room $i$ only and flow rate are used to identify $\hat{\Theta}^{T}_{\tT^i_j}$ for temperature models in $S_2$. All the models have been trained on the data from March $11^{th}$ $2016$ to April $26^{th}$ $2016$ and validated on the data from May $1^{st}$ $2016$ to May $15^{th}$ $2016$. The accuracy of these models with respect to real data is shown in Table \ref{T:S1accuracy}, based on the definition of Normalized Root Mean Square Error (NRMSE).

@@

%\begin{figure}[h!]
%	\begin{center}
%		\includegraphics[width=26pc]{figures/Energy_all.eps}
%	\end{center}
%	\caption{Comparison of DPC and bang-bang control performance with different violations using random forest model.}
%	\label{F:comparison_all_energy}
%\end{figure}
%An alternative comparison can be done considering the EnergyPlus model of the house for energy consumption described in Section \ref{SS:energyPlusmodel}. Results are shown in Figure \ref{F:comparison_all_energy_E+}.
In Figure \ref{F:comparison_all_energy_E+}, using the thermal energy consumption model derived in Section \ref{SS:energyPlusmodel} using EnergyPlus, we show how DPC outperforms the bang-bang controller also in terms of energy consumption and how the bounds violations allow us to save more energy. In this case, since the plot is clear and we are interested in showing energy saving on a long period, we ran the simulations over the testing data, i.e. $15$ days of May, from May $1^{st}$ $2016$ to May $15^{th}$ $2016$. We can see that if we want to keep the temperature within a comfort range, only allowing small violations, the use of DPC produce an energy saving of $45\, kWh$, that correspond to the $25.4\%$ over a period of $15$ days. Instead if we allow a large violation over the same period, the energy saving is of $87\, kWh$, that correspond to the $49.2\%$.
\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=26pc]{figures/Energy_all_EnergyPlus.eps}
	\end{center}
	\caption{Comparison of DPC and bang-bang controller performance in terms of thermal energy saving with different violations using EnergyPlus model.}
	\label{F:comparison_all_energy_E+}
\end{figure}

>>>>

\section{REAL CASE STUDY: OPTIMAL HEATING SYSTEM SCHEDULING}
\label{S:realCaseStudy}

In this section we show an application of DPC to a real case study. In particular, in the following we first describe the plant structure, then we provide both data-based and EnergyPlus models of such a plant to be used for control and performance comparison, and finally we show results of the closed-loop simulations.

@@

\begin{figure}[h!]
	\begin{center}
		\subfigure[Walls.]{
			\label{F:houseSection1}
			\centering
			\includegraphics[width=20pc]{figures/blocco_parete_rev01.eps}
		}
		\subfigure[Floor and roof.]{
			\label{F:houseSection2}
			\centering
			\includegraphics[width=20pc]{figures/blocco_solaio_rev01.eps}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{Cross-sections of the wood-cement blocks.}
	\captionsetup{justification=centering}
	\label{F:houseSection}
\end{figure}
The heating system of the building, that supplies the thermal energy required in winter season, is a hydronic system consisting of a vegetable biomass boiler, with a manual ON/OFF, a constant-flow pump station, and tubular steel radiators as shown in Figure \ref{F:housePlantScheme}. The standard biomass boiler has an efficiency of $83.5\%$ with $16.5\ kW$ of thermal power transferred to the water, without gas-flame modulation. The heat transfer fluid distribution is realized through a manifold circuit, that supplies the radiators placed in the various rooms of the house. The energy needs for the domestic hot water (DHW) are covered by the same boiler, coupled with a solar thermal plant. It is worth noting that, in this work, the thermal energy needed for the domestic hot water production is neglected.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=27pc]{figures/Schema_di_impianto.eps}
		\caption{Technological plant scheme of the use case.}
		\captionsetup{justification=centering}
		\label{F:housePlantScheme}
	\end{center}
\end{figure}
For a thorough knowledge of the building thermal behavior, an in-situ analysis was carried out. Temperature measuring devices, completely self-produced at the “G. Parolini Lab” of the University of L'Aquila \cite{Pantoli2017}, mainly based on an $ATmega 2560$ microcontroller and $DS18B20$ temperature probes (temperature range from $-55.0\degree C$ to $125.0\degree C$ with an accuracy of $\pm 0.5\degree C$), were employed to acquire the ambient temperatures of the four differently oriented rooms of the house as shown in Figure \ref{F:houseFloors}. The probes positions were chosen to optimize the data acquisition and to minimize the discomfort of the occupants in the house. Furthermore, as can be seen in Figures \ref{F:housePlantScheme} and \ref{F:houseFloors}, a commercial heat meter (temperature range from $10.0\degree C$ to $90.0\degree C$ with an accuracy of $\pm 0.05\degree C$) was installed downstream of the biomass boiler to measure the produced thermal energy. The turbine flowmeter was installed on the return pipe, while the two thermocouples were placed inside the delivery and return pipes, respectively. All the measuring devices have been set with a data acquisition rate equal to $10$ minutes, from March $11^{th}$ $2016$ to May $15^{th}$ $2016$.
\begin{figure}[h!]
	\begin{center}
		\subfigure[Ground floor.]{
			\label{F:houseGroundFloor}
			\centering
			\includegraphics[width=26pc]{figures/Arch_PT_e_sonde.eps}
		}
		\subfigure[Attic.]{
			\label{F:houseAttic}
			\centering
			\hspace{-1.6cm}
			\includegraphics[width=22.8pc]{figures/Arch_P1.eps}
		}
	\end{center}
	\caption{Layout of the house and probes placement. Legend: red circle for ambient temperature; orange circle for heat meter.}
	\captionsetup{justification=centering}
	\label{F:houseFloors}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=20pc]{figures/Render_rev01.eps}
		\caption{Virtual model of the use case.}
		\captionsetup{justification=centering}
		\label{F:houseVirtualModel}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\subfigure[Thermal energy.]{
			\label{F:houseComparisonExperimentalEnergy}
			\centering
			\includegraphics[width=24pc]{figures/Thermal_energy_rev01.eps}
		}
		\subfigure[Thermal power.]{
			\label{F:houseComparisonExperimentalPower}
			\centering
			\includegraphics[width=24pc]{figures/Thermal_power.eps}
		}
	\end{center}
	\caption{Comparison between numerical and experimental data.}
	\captionsetup{justification=centering}
	\label{F:houseComparisonExperimental}
\end{figure}
\subsection{Random forest models} For the closed-loop simulations with DPC we create $2$ different sets of models, $S_1 = \{\tP^{r}_{\mathrm{k+1|k}},\tT^{r_1}_{\mathrm{k+1|k}},\tT^{r_2}_{\mathrm{k+1|k}},\tT^{r_3}_{\mathrm{k+1|k}},\tT^{r_4}_{\mathrm{k+j|k}}\}$ and $S_2 = \{\tP_{\mathrm{k+j|k}},\tT^{1}_{\mathrm{k+j|k}},\tT^{2}_{\mathrm{k+j|k}},\\\tT^{3}_{\mathrm{k+j|k}},\tT^{4}_{\mathrm{k+j|k}},\ j = 1,\ldots,N\}$, using random forests. In each set we have a model for the power consumption of the house ($\tP^{r}$ and $\tP$) and $4$ models that describe the evolution of the room temperature in each of the $4$ rooms ($\tT^{r_i}$ and $\tT^i$). Models in $S_1$ are created using the classical random forests algorithm with all the features and are used as plant simulator of the house. For this reason, they are computed to give prediction only for time step $k+1$ and not over the whole horizon $N$. Models in $S_2$ are used as predictors over a horizon $N$ in the DPC algorithm and are then created using the methodology provided in Section \ref{SS:dpcrf}. The non-manipulated features in $\tX^d$ are the disturbance data (relative humidity, atmospheric pressure, outside air temperature, solar radiation, wind, time of the day and day of the week) and the states (temperature of the $4$ rooms). The manipulated feature in $\tX^c$ is the flow rate ($[m^3/h]$). All this features are used to create models in $S_1$ and the power model in $S_2$, while disturbance data, state temperature of room $i$ only and flow rate are used to identify $\hat{\Theta}^{T}_{\tT^i_j}$ for temperature models in $S_2$. All the models have been trained on the data from March $11^{th}$ $2016$ to April $26^{th}$ $2016$ and validated on the data from May $1^{st}$ $2016$ to May $15^{th}$ $2016$. The accuracy of these models with respect to real data is shown in Table \ref{T:S1accuracy}, based on the definition of Normalized Root Mean Square Error (NRMSE).

@@

%\begin{figure}[h!]
%	\begin{center}
%		\includegraphics[width=26pc]{figures/Energy_all.eps}
%	\end{center}
%	\caption{Comparison of DPC and bang-bang control performance with different violations using random forest model.}
%	\label{F:comparison_all_energy}
%\end{figure}
%An alternative comparison can be done considering the EnergyPlus model of the house for energy consumption described in Section \ref{SS:energyPlusmodel}. Results are shown in Figure \ref{F:comparison_all_energy_E+}.
In Figure \ref{F:comparison_all_energy_E+}, using the thermal energy consumption model derived in Section \ref{SS:energyPlusmodel} using EnergyPlus, we show how DPC outperforms the bang-bang controller also in terms of energy consumption and how the bounds violations allow us to save more energy. In this case, since the plot is clear and we are interested in showing energy saving on a long period, we ran the simulations over the testing data, i.e. $15$ days of May, from May $1^{st}$ $2016$ to May $15^{th}$ $2016$. We can see that if we want to keep the temperature within a comfort range, only allowing small violations, the use of DPC produce an energy saving of $45\, kWh$, that correspond to the $25.4\%$ over a period of $15$ days. In case of medium violations get an energy saving of $57\, kWh$, that is $32.2\%$. Instead if we allow a large violation over the same period, the energy saving is of $87\, kWh$, that correspond to the $49.2\%$.
\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=26pc]{figures/Energy_all_EnergyPlus.eps}
	\end{center}
	\caption{Comparison of DPC and bang-bang controller performance in terms of thermal energy saving with different violations using EnergyPlus model.}
	\label{F:comparison_all_energy_E+}
\end{figure}

### 2017-09-26 paper/intro.tex
Modified introduction. First complete draft.

Consider a multivariable dynamical system subject to external disturbances. The first and foremost requirement for making any decision is to obtain the underlying control-oriented predictive model of the system. With a reasonable forecast of the external disturbances, these models should predict the state of the system in the future and thus a predictive controller based on Model Predictive Control (MPC) can act preemptively to provide a desired behavior. In particular, MPC has been proven to be very powerful for multivariable systems in the presence of input and output constraints, and forecast of the disturbances. The caveat is that MPC requires a reasonably accurate physical representation of the system. This makes MPC unsuitable for control of complex plants such as natural gas processing, oil refineries, boilers, manufacturing plants, and buildings where the user expertise, time, and associated sensor costs required to develop a model are very high \cite{Sturzenegger2016,vzavcekova2014}.

@@

The question now is, can we employ data-driven techniques to reduce the cost of modeling, and still exploit the benefits that MPC has to offer? We therefore look for automatic and data-driven approaches to control that are also adaptive, scalable and interpretable. We solve this problem with \textit{Data Predictive Control (DPC)} by bridging the gap between Machine Learning and Predictive Control.

In our previous work~\cite{Behl2016,Jain2016,JainACC2017,JainCDC2017}, we introduced the concept of DPC for receding horizon control.
This work has the following contributions.
(1) We first formally present two underlying algorithms: (i) DPC with regression trees, and (ii) DPC with random forests.
(2) Using a bilinear building model whose parameters were identified using experiments on a building in Switzerland, we demonstrate the strength of DPC for receding horizon control via one-to-one comparison against a benchmark MPC controller. We show DPC captures 70\% variance in MPC and offers a comparable performance.
(3) We present a practical application of DPC for Demand Response, where we apply DPC to a 6 story 22 zone building model in EnergyPlus \cite{Crawley2001} for which model-based control is not economical and practical due to extreme complexity. We show scalability and efficiency of DPC in providing financial incentives to the end-customers bypassing the need for high fidelity models. We observe that DPC provides the desired power reduction with an average error of 3\%.

>>>>

Consider a multivariable dynamical system subject to external disturbances. The first and foremost requirement for making any decision is to obtain the underlying control-oriented predictive model of the system. With a reasonable forecast of the external disturbances, these models should predict the state of the system in the future and thus Model Predictive Control (MPC) can act preemptively to provide a desired system behavior while optimizing a desired performance. In particular, MPC has been proven to be very powerful for multivariable systems in the presence of input and output constraints, and forecast of the disturbances. The caveat is that MPC requires a reasonably accurate physical representation of the system. This makes MPC unsuitable for control of complex plants such as natural gas processing, oil refineries, boilers, manufacturing plants, and buildings where the user expertise, time, and associated sensor costs required to develop a model are very high \cite{Sturzenegger2016,vzavcekova2014}.

@@

The question now is, can we employ data-driven techniques to reduce the cost of modeling and still exploit the benefits that MPC has to offer? Can we build automatic and data-driven approaches for control purposes that are also adaptive, scalable and interpretable? In this paper we address this problem introducing \textit{Data Predictive Control (DPC)} to bridge the gap between Machine Learning and Predictive Control.

In literature several studies can be found dealing with data-based control techniques that make use of machine learning algorithms \cite{Hou2013}. For example in \cite{Afram2017} Artificial Neural Networks are exploited to create data-based system models to be used as plant simulator in closed-loop with a Suervisory Model Predictive Control. In \cite{Macarulla2017} the authors proposed a predictive control strategy based on Neural Networks, for boilers control in buildings, to decide the optimal time to switch the plan on to guarantee energy saving and thermal comfort. However the strategy is not easily scalable to different types of plants and does not use optimization theory in the closed-loop scheme. In \cite{Costanzo2016} a reinforcement learning control strategy, called Model-Assisted Batch Reinforcement Learning, is considered to provide data-driven control for the demand response problem in HVAC systems. Reinforcement Learning is a model-free methodology and for this reason it can not be used for on-line optimization in a closed-loop predictive control scheme. In \cite{Ferreira2012} the authors considered a data-driven predictive control based on Neural Networks. They use this approach to guarantee energy saving and  thermal comfort in public buildings. Neural Networks are used in the closed-loop control scheme to determine a thermal comfort index based on parameters that can be measured or estimated, but no  system dynamics with internal state are included into the optimization problem. More papers related to this topic can be found in literature, but to the best of the authors knowledge none of them address the problem of including data-based state models in the optimal predictive control loop, and hence allow to set up a MPC-like optimal control problem, as we do in DPC.

\paragraph{Main contributions and paper organization}
In our previous work~\cite{Behl2016,Jain2016,JainACC2017,JainCDC2017}, we introduced the concept of DPC for receding horizon control. In this paper we extend these results providing the following contributions:
\begin{enumerate}
	\item in Section \ref{S:dpc} we formally present the two following control techniques:
	\begin{itemize}
		\item DPC with regression trees;
		\item DPC with random forests;
	\end{itemize}
	\item in Section \ref{S:proof}, using a bilinear building model whose parameters were identified using experiments on a building in Switzerland, we demonstrate the strength of DPC for receding horizon control via one-to-one comparison against a benchmark MPC controller. We show DPC captures 70\% variance in MPC and offers a comparable performance;
	\item in Section \ref{S:casestudy}, we present a practical application of DPC for Demand Response, where we apply DPC to a 6 story 22 zone building model in EnergyPlus \cite{Crawley2001} for which model-based control is not economical and practical due to extreme complexity. We show scalability and efficiency of DPC in providing financial incentives to the end-customers bypassing the need for high fidelity models. We observe that DPC provides the desired power reduction with an average error of 3\%;
	\item in Section \ref{S:realCaseStudy}, we implement DPC on real data coming from an off-grid house located in L'Aquila (Italy), to find the optimal ON/OFF scheduling for the heating system in order to save energy while guaranteeing thermal comfort for the occupants. We quantify the total amount of energy saved, with respect to the classical bang-bang controller widely used in houses for temperature control, using an EnergyPlus model built specifically for the house. We show we can perform an energy saving that goes from $25\%$, if we guarantee thermal comfort, to $50\%$, if we allow very little discomfort in terms of rooms temperature.
\end{enumerate}
